
# Libraries

R comes with a bit of functionality. However, most of the useful tools in R is distributed as packages. There are +10.000 package for R, so it is a jungle to figure out what the most easy solution to your problem at hand is. However, the teams who have made tidyverse and ggplot2 etc. have made a lot of things much more easy, and we strongly rely on their tools and routines in data analysis. 

To install packages from CRAN (the main repo where R-packages are distributed)

```{r, eval=FALSE}
install.packages('somepackage')
```

To install packages from github (the place where all the development and general code sharing is distributed)

```{r,eval=FALSE}
devtools::install_github('developername/packagename')
```

To make packages available within your analysis use library(), or use the package name followed by :: and the function

```{r}
library(ggplot2) # lets plot dadta
library(tidyverse)
library(ggpubr) # lets add stats to the plots
library(knitr) # lets make nice tables

ggplot2::qplot(rnorm(100)) # example of a function call without library'ing the package.
```




# Descriptive statistics and plotting

```{r,eval=FALSE, include=TRUE}
load('iBuffetSurveyData.RData')
```

```{r,eval=T, include=F}
load('~/Dropbox/Backup/MyDocumentsOnC/Course and teaching/Meal Systems and Technology/data/BuffetData2/BuffetSurveyData.RData')
```

## Distributions of count data

The table function is good in getting how many observations there are within a given vector, or combinations of several vectors.

Here used on the Survey answers. 
```{r}
table(Surveylong$answ) # across everything
table(Surveylong$question,Surveylong$answ ) # across question
```
You see that most of the answers are in agreement with question, and that there are no observations in the _Strongly disagree_ category. 

_The tidyverse way_ 

Lets do exactly the same just using tidyverse functions count(), group_by(), mutate(), and  summarise().

Further, lets print the results in a nice looking table using kable() from the knitr package. 

```{r}
# tb <- Surveylong %>% 
  # count(question,answ,Day,name = "no_rows", .drop = F) 

# kable(tb, caption = 'some caption')
```

The numbers are absolute, but may be better represented by proportions. 

```{r}
tb <- Surveylong %>% 
  group_by(question,Day) %>% 
  dplyr::mutate(ntot = n()) %>% 
  group_by(question,answ,Day) %>% 
  dplyr::summarise(n = n(), 
            prc = 100*n / ntot[1])

kable(tb, caption = 'some caption', digits = 1)
```

... and a plot of it

```{r}
tb %>% 
  ggplot(data = ., aes(answ,prc, fill = factor(Day))) + 
  geom_bar(stat = 'identity', position = position_dodge()) + 
  facet_wrap(~question) + 
  theme(axis.text.x = element_text(angle = 45,hjust = 1), 
        legend.position = 'top')
```

### Descriptives for a continouos variable

Here just across the entire sample set. 

```{r}
mean(Buffet_survey$Consumption)
median(Buffet_survey$Consumption)
sd(Buffet_survey$Consumption)
IQR(Buffet_survey$Consumption)
summary(Buffet_survey$Consumption)
```


This is a very high level representation, and we usually want to compare means (or other metrics) between different groups. 

We use the consumption and split it according to day and pasta-type.

```{r}
tb2 <- Buffet_survey %>% 
  group_by(StationName, Day) %>% 
  dplyr::summarise(nobs = n(),
            mean = mean(Consumption), 
            median = median(Consumption),
            sd = sd(Consumption),
            iqr = IQR(Consumption), 
            q25 = quantile(Consumption,0.25), 
            q75 = quantile(Consumption,0.75))

kable(tb2, digits = 1, caption = 'some relevant caption')
```

Corresponding plot of data 

```{r}
Buffet_survey %>% 
  ggplot(data = ., aes(StationName,Consumption, fill = factor(Day))) + 
  geom_violin()

Buffet_survey %>% 
  ggplot(data = ., aes(StationName,Consumption, fill = factor(Day))) + 
  geom_boxplot() +
  geom_jitter()

```

Corresponding plots of the results

```{r}
ggplot(data= tb2, aes(factor(StationName):factor(Day),mean, 
                      color = factor(Day), ymin = mean-sd, ymax = mean + sd)) +
  geom_point() + 
  geom_errorbar(width = 0.3)+
  theme(axis.text.x = element_text(angle = 45,hjust = 1)) + 
  ylab('mean+/- 1*sd')
```

#### In relation to protein considerations

Try to make these descriptive analysis and plots taking into account whether the participants considered protein content, and why they did.

Here is some inspiration. 

```{r}
Buffet_survey %>% 
  ggplot(data = ., 
         aes(`Did you consider the protein content of the dish(es) you chose?`,
             Consumption)) + 
  geom_boxplot() + 
  geom_jitter() + 
  facet_wrap(~StationName)

```


## Scatter plots

Lets plot the consumption as a function of the answers to the liking-scale questions of the survey, and split it into day and type of vegetable. 
If you think about it, it is pretty many plots, but the ggplot2 functionality facet_wrap() on a long format data frame does it in few lines: 


```{r, fig.height=9}
Surveylong_buffet %>% 
  filter(!is.na(StationName )) %>% 
  mutate(question2 = question %>% substr(1,34)) %>% # The label is to long, so lets just represent the first 30 letters.
  ggplot(data = .,  aes(answnum,Consumption, color = factor(Day))) +
  geom_point() + 
  stat_smooth(se = F, method = lm) + 
  stat_cor() +
  facet_grid(question2 ~ StationName) + 
  theme_bw() + theme(legend.position = 'bottom')
```

```{r}
kable(Surveyscales, caption = 'Just a table to have what the 7point likert scale numbers mean')
```

Get this stuff to work, and try to interpret what you see: which factors seems important for the portion size ( _Consumption_ )? 

The summary-stats in the scatter plots above: What is this? How is it interpreted? 

  