
# Introduction to PCA

In this book Principal Component Analysis (PCA) is used several times. This chapter will shortly explain the theory behind PCA and the interpretation of relevant plots.  

PCA is a tool for looking a correlation structure between variables, and groupings of samples. All through visualizations.  
Check out youtube on the subject for an introduction.

[TILFØJ: skrives kort om PCA, skal der ikke stå lidt mere???]


```{r}
knitr::include_url("https://youtube.com/embed/NFIkD9-MuTY")
```

![](https://youtu.be/NFIkD9-MuTY){width="90%"}


[INDSÆTTE: flere film // you tube klip]



# Introduction linear and mixed models 
[ALT MANGLER - forslag er at lave et kortere afsnit her som en pendant til PCA. Håber I synes det er en god ide?]





# Libraries

R comes with a bit of functionality. However, most of the useful tools in R is distributed as packages. There are +10.000 package for R, so it is a jungle to figure out what the most easy solution to your problem at hand is. However, the teams who have made tidyverse and ggplot2 etc. have made a lot of things much more easy, and we strongly rely on their tools and routines in data analysis. 

To install packages from CRAN (the main repo where R-packages are distributed)

```{r, eval=FALSE}
install.packages('somepackagename')
```

To install packages from github (the online place where all the development and general code sharing is distributed)

```{r,eval=FALSE}
devtools::install_github('developername/somepackagename')
```

To make packages available within your analysis use library(), or use the package name followed by :: and the function. The library function will activate the installed package. 

```{r}
library(ggplot2) # lets plot dadta
library(tidyverse)
library(ggpubr) # lets add stats to the plots
library(knitr) # lets make nice tables

ggplot2::qplot(rnorm(100)) # example of a function call without library'ing the package.
```

[MANGLER: info på tidyverse]




# Descriptive statistics
In this Chapter we will go through the main elements you need to 


[MANGLER: bla bla bla introtekst.... inklu ]


We are using the combined buffet consumption and survey dataset. We have already made a merged version for you, so yo just need to load the data. 


## Loading data

First we need to load the data. If you do not have a saved data set (*.RData), please go to the chapters about "How to import data" and "How to save the data" for more info on import and data saving. 

Use this if your data is in the same folder as your script: 
```{r,eval=FALSE, include=TRUE}
load('iBuffetSurveyData.RData')
```

Use this, if you need to browse for your *.RData file: 
```{r,eval=T, include=F}
load('~/Dropbox/Backup/MyDocumentsOnC/Course and teaching/Meal Systems and Technology/data/BuffetData2/BuffetSurveyData.RData')
```

[MORTEN: skal vi stadig kalde RData filen dette her? Giver det mening?]


## Distributions of count data

The table function is good in getting how many observations there are within a given vector, or combinations of several vectors.

Here used on the Survey answers. 
```{r}
table(Surveylong$answ) # across everything
table(Surveylong$question,Surveylong$answ ) # across question
```
You see that most of the answers are in agreement with question, and that there are no observations in the _Strongly disagree_ category. 

[MORTEN: kan vi lave en anden form for tabel her? Meld gerne ind hvad type data det kunne være?? for den man får ud giver ikke så meget mening set fra vores side. og så skal vi overveje om "tidyverse" er nice to know viden?]



[MÅSKE: The nice to know part:]

_The tidyverse way_ 

Lets do exactly the same just using tidyverse functions count(), group_by(), mutate(), and  summarise().

Further, lets print the results in a nice looking table using kable() from the knitr package. 

```{r}
# tb <- Surveylong %>% 
  # count(question,answ,Day,name = "no_rows", .drop = F) 

# kable(tb, caption = 'some caption')
```



The numbers are absolute, but may be better represented by proportions. 

```{r}
tb <- Surveylong %>% 
  group_by(question,Day) %>% 
  dplyr::mutate(ntot = n()) %>% 
  group_by(question,answ,Day) %>% 
  dplyr::summarise(n = n(), 
            prc = 100*n / ntot[1])

kable(tb, caption = 'some caption', digits = 1)
```

... and a plot of it

```{r}
tb %>% 
  ggplot(data = ., aes(answ,prc, fill = factor(Day))) + 
  geom_bar(stat = 'identity', position = position_dodge()) + 
  facet_wrap(~question) + 
  theme(axis.text.x = element_text(angle = 45,hjust = 1), 
        legend.position = 'top')
```


### Descriptives for a continouos variable

[MORTEN: skal denne også ændres til noget smartere? datasættet er opsæt anderledes end vi normalt ville gøre det, og jeg tænker mere og mere på, om det er hensigtsmæssigt? Hjææælp...]

Here just across the entire sample set. 

```{r}
mean(Buffet_survey$Consumption)
median(Buffet_survey$Consumption)
sd(Buffet_survey$Consumption)
IQR(Buffet_survey$Consumption)
summary(Buffet_survey$Consumption)
```


This is a very high level representation, and we usually want to compare means (or other metrics) between different groups. 

We use the consumption and split it according to day and pasta-type.

```{r}
tb2 <- Buffet_survey %>% 
  group_by(StationName, Day) %>% 
  dplyr::summarise(nobs = n(),
            mean = mean(Consumption), 
            median = median(Consumption),
            sd = sd(Consumption),
            iqr = IQR(Consumption), 
            q25 = quantile(Consumption,0.25), 
            q75 = quantile(Consumption,0.75))

kable(tb2, digits = 1, caption = 'some relevant caption')
```



# Plotting data

Corresponding plot of data 

```{r}
Buffet_survey %>% 
  ggplot(data = ., aes(StationName,Consumption, fill = factor(Day))) + 
  geom_violin()

Buffet_survey %>% 
  ggplot(data = ., aes(StationName,Consumption, fill = factor(Day))) + 
  geom_boxplot() +
  geom_jitter()

```

Corresponding plots of the results

```{r}
ggplot(data= tb2, aes(factor(StationName):factor(Day),mean, 
                      color = factor(Day), ymin = mean-sd, ymax = mean + sd)) +
  geom_point() + 
  geom_errorbar(width = 0.3)+
  theme(axis.text.x = element_text(angle = 45,hjust = 1)) + 
  ylab('mean+/- 1*sd')
```

#### In relation to protein considerations

Try to make these descriptive analysis and plots taking into account whether the participants considered protein content, and why they did.

Here is some inspiration. 

```{r}
Buffet_survey %>% 
  ggplot(data = ., 
         aes(`Did you consider the protein content of the dish(es) you chose?`,
             Consumption)) + 
  geom_boxplot() + 
  geom_jitter() + 
  facet_wrap(~StationName)

```


## Scatter plots

Lets plot the consumption as a function of the answers to the liking-scale questions of the survey, and split it into day and type of vegetable. 
If you think about it, it is pretty many plots, but the ggplot2 functionality facet_wrap() on a long format data frame does it in few lines: 

```{r, fig.height=9}
Surveylong_buffet %>% 
  filter(!is.na(StationName )) %>% 
  mutate(question2 = question %>% substr(1,34)) %>% # The label is to long, so lets just represent the first 30 letters.
  ggplot(data = .,  aes(answnum,Consumption, color = factor(Day))) +
  geom_point() + 
  stat_smooth(se = F, method = lm) + 
  stat_cor() +
  facet_grid(question2 ~ StationName) + 
  theme_bw() + theme(legend.position = 'bottom')
```


## How to export plots 

The plots shown in the _Plots_ pane (lower right of RStudio) can be saved using the _Export_ bottom. 

You can also save the plots to a file using the ggsave() function. 
If you run ggsave() without specifying which plot to export, it will use the latest. 
You can also directly specify the plot to export: 

```{r, fig.height=9, eval = F}
myplot <- Surveylong_buffet %>% 
  filter(!is.na(StationName )) %>% 
  mutate(question2 = question %>% substr(1,34)) %>% # The label is to long, so lets just represent the first 30 letters.
  ggplot(data = .,  aes(answnum,Consumption, color = factor(Day))) +
  geom_point() + 
  stat_smooth(se = F, method = lm) + 
  stat_cor() +
  facet_grid(question2 ~ StationName) + 
  theme_bw() + theme(legend.position = 'bottom')

ggsave('liking_vs_consumption.pdf',myplot)

```

ggsave() supports different formats (.png, .tiff, .pdf,...) and further allows for editing the size (height = , width = ). 

```{r}
kable(Surveyscales, caption = 'Just a table to have what the 7point likert scale numbers mean')
```

Get this stuff to work, and try to interpret what you see: which factors seems important for the portion size ( _Consumption_ )? 

The summary-stats in the scatter plots above: What is this? How is it interpreted? 
