
# Linear models

Linear models is a general term for models with a single univariate response (dependent variable - $y$ in the formula below), which we want to describe using one or several predictors (independent variables - $x$ in the formula below). 

$$  y = a + b \cdot x + e $$
Here the informative parameter is the slope ($b$) which indicates the relation between $x$ and $y$. ($e$ is the missfit / residuals of the model). 


We use tidyverse coding as this makes life much easier. As an tidyverse add on, we use broom for the linear models, broom.mixed and lme4 for the linear mixed models.


```{r}
library(ggplot2) # lets plot dadta
library(tidyverse)
library(broom)
library(broom.mixed)
library(lme4)
library(ggpubr) # lets add stats to the plots
library(knitr) # lets make nice tables
```

The data is already imported, and formated (see _Getting_data_in.pdf_ for details). We simply load this file. 

```{r}
library(data4consumerscience)
```



## Example

As response variable, the amount of Consumption of _Pasta with mushrooms_ and use the likert scale _`I like the taste of pasta with mushrooms!`_ as predictor. We use ONLY Day 1 results.  

First a plot:

```{r}
pasta %>% 
  filter(str_detect(StationName,'mush')) %>% 
  filter(Day==1) %>% 
  ggplot(data = ., aes(I_like_taste_of_pasta_with_mushrooms,Consumption)) + 
  geom_point() + 
  stat_smooth(method = lm, se = F)
```

It seems as there is something. 
So lets build a linear model on this 

```{r}
# subset the data
x <- pasta %>% 
  filter(str_detect(StationName,'mush')) %>% 
  filter(Day==1) 

mdl <- lm(data = x, Consumption~I_like_taste_of_pasta_with_mushrooms) 
mdl  
summary(mdl)
```

The the slope indicates that by increasing liking by one unit the consumption increase is $50.2 \pm 30.4$, however, this apparent effect is not statistically significant ($p = 0.12$). 

## Run a bunch of models at once

We want to model consumption of both pasta with mushrooms andd legumes, and look at all the likert scales questions as predictors. Further we want to do this for both days. 

First we create a new long format data frame

```{r}
pastalong <- pasta %>% 
  gather(question,answ,I_like_taste_of_pasta_with_legumes:Pasta_with_mushrooms_is_visually_appealing) %>% 
  mutate(answnum = factor(answ,labels = c('Disagree','More or less disagree','Neither agree nor disagree','More or less agree','Agree','Strongly agree')) %>% as.numeric())
```


### A plot 

```{r, fig.height=9}
pastalong %>% 
  filter(!is.na(StationName )) %>% 
  mutate(question2 = question %>% substr(1,34)) %>% # The label is to long, so lets just represent the first 30 letters.
  ggplot(data = .,  aes(answnum,Consumption, color = factor(Day))) +
  geom_point() + 
  stat_smooth(se = F, method = lm) + 
  stat_cor() +
  facet_grid(question2 ~ StationName) + 
  theme_bw() + theme(legend.position = 'bottom')
```

This we similary can run as several linear models. 



```{r}
tb <- pastalong %>% 
  filter(!is.na(StationName )) %>% 
  group_by(StationName,question,Day) %>% 
  do(lm(data = ., Consumption~answnum) %>% tidy(conf.int = T))

tb %>% 
  filter(term=='answnum') %>% 
  select(-statistic) %>% 
kable(x = .,caption = 'All linear models', digits = 2)

```
.. A plot of these results for a quick interpretation. 

```{r}
tb %>% 
  filter(term=='answnum') %>% 
  ggplot(data = ., aes(question,estimate,ymin = conf.low, ymax = conf.high, color = factor(Day))) + 
  geom_errorbar(width = 0.1, position = position_dodge()) +geom_point()+ 
  geom_hline(yintercept = 0) + 
  coord_flip() +facet_grid(~StationName) + 
  theme(legend.position = 'bottom')
```

Seems as some of the legume consumptions there is a significant association with likert scales. Not as strong for consumption of mushrooms.  

# Mixed models

Mixed models are used when there is repetitions in the response due to (here) the person conducting the trial. 
The two days are repetitions, and hence we can use all the data (not splitting in to days), but need to account for the person in the model.  


```{r}
# subset the data
x <- pasta %>% 
  filter(str_detect(StationName,'mush')) 

mdl <- lmer(data = x, Consumption~I_like_taste_of_pasta_with_mushrooms + Day +  (1|Person)) 
summary(mdl)
```

This is the joint effect between the two days. Think of an average of the two slopes - one for each day -. Here taking into account that each person has provided two responses of the consumption of pasta with mushrooms. 

This can also be accomplised using the tidyverse setup engined by the broom.mixed package.

In principle, we simply do not loop over Day, but include it in the formula along with person.


```{r}
tbmixed <- pastalong %>% 
  filter(!is.na(StationName )) %>% 
  group_by(StationName,question) %>% 
  do(lmer(data = ., Consumption~answnum + Day + (1|Person)) %>% tidy(conf.int = T))
```

The output here is a bit different than the lm() model. But it is still the slope of answnum which carries the interesting stuff. 

```{r}
tbmixed %>% 
  filter(term=='answnum') %>% 
  select(-effect,-group) %>% 
kable(x = .,caption = 'All mixed linear models', digits = 2)

tbmixed %>% 
  filter(term=='answnum') %>% 
  ggplot(data = ., aes(question,estimate,ymin = conf.low, ymax = conf.high)) + 
  geom_errorbar(width = 0.1) +geom_point()+ 
  geom_hline(yintercept = 0) + 
  coord_flip() +facet_grid(~StationName) + 
  theme(legend.position = 'bottom')
```

Do the associations match as expected? 

## With several variables

We can add several predictors to the model, here that could several likert-scale questions, and maybe demographics with the consumption as response. 
This is in principle the same for both linear models and linear mixed models. 

```{r}
x <- pasta %>% 
  filter(str_detect(StationName,'mush'))

mdl <- lmer(data = x, Consumption~I_like_taste_of_pasta_with_mushrooms + 
       Pasta_with_mushrooms_is_visually_appealing + Day + (1|Person))

summary(mdl)
mdl %>% tidy(conf.int = T)
```

Try to interpret the slopes? Are the slopes significantly different from 0 (i.e. the point of no association). .. And hey! Why is the slope for visual all of a sudden negative?... Does that mean that consumption increase the less you like the visual appearance? .. Or what?  

Complete the same analysis with legumes. 

```{r, eval=F, include=F}
x <- pasta %>% 
  filter(str_detect(StationName,'legu'))

mdl <- lmer(data = x, Consumption~I_like_taste_of_pasta_with_legumes +
              Pasta_with_legumes_is_visually_appealing + Day + (1|Person))

summary(mdl)
mdl %>% tidy(conf.int = T)
```
